A Responsible Machine Learning Workflow\\\normalsize{With Focus on Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2020-01-31 article-information-2019.tex
first pass on XNN sim results

\noindent where $\mathbf{X}_{\text{Friedman},j}$ are random uniform features in $[0,1]$. In Friedman's texts, a Gaussian noise term was added to create a continuous output feature for testing spline regression methodologies. In this manuscript, the signal generating function and input features are modified in several ways.  Two binary features, a categorical feature with five discrete levels, and a bias term are introduced into $f$ to add a degree of complexity that may more closely mimic real-world settings. 
%When generating the data, random draws for each observation are constructed such that the features can be correlated with each other, some highly so. For example, X1 and X3 have a Pearson correlation coefficient equal to 0.58.
For binary classification analysis, the Gaussian noise term is replaced with noise drawn from a logistic distribution and coefficients are re-scaled to be one fifth of the size of those used by Friedman, and any $f(\mathbf{X})$ value above 0 is classified as a positive outcome, while $f(\mathbf{X})$ values less than or equal to zero are designated as negative outcomes.  Finally, $f$ is augmented with two hypothetical protected class-control features with known dependencies on the binary outcome to allow for discrimination testing. The simulated data is generated to have eight input features, twelve after numeric encoding of categorical features, and a binary outcome, two class-control features, and 100,000 rows. The simulated data is then split into a training and test set, with 80,000 and 20,000 observations, respectively.  Within the training set, a 5-fold cross-validation indicator is used for training all models. For an exact specification of the simulated data, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints, with respect to each input feature, $X_j$, and a target feature, $\mathbf{y}$, independently. An MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, and each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$. For some row of data, $\mathbf{x}$, a trained MGBM model, $g^{\text{MGBM}}$, takes the form:

@@

\noindent As in unconstrained GBM, $\Theta^{\text{mono}}_b$ is selected in a greedy, additive fashion by minimizing a regularized loss function that considers known target labels, the predictions of all subsequently trained trees in $g^{\text{MGBM}}$, and the $b$-th tree splits applied to $\mathbf{x}$, $T_b(\mathbf{x}^{(i)};\Theta^\text{mono}_b)$, in a numeric error function (e.g., squared error, Huber error), and a regularization term that penalizes complexity in the current tree. See Appendices \ref{a_ssec:not} and \ref{a_ssec:mgbm} for details pertaining to MGBM training.

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bear a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e., GAMs which consider main effects and a small number of 2-way interactions and may also incorporate boosting into their training \cite{esl}, \cite{ga2m}.  Hence, XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. A trained XNN function, $g^{\text{XNN}}$, applied to some row of data, $\mathbf{x}$, is defined as:

@@

Two $g^\text{XNN}$ models are trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in 5 training folds and in a test data partition. $L_1$ regularization is applied to both the lowest projection layer and to the $n_k$ subnetworks to induce a sparse and interpretable model, where each $n_k$ and corresponding $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft} and Appendices  \ref{a_ssec:not} and \ref{a_ssec:xnn}.

@@

PD plots are a widely-used method for describing and plotting the estimated average predictions of a complex model $g$ across some partition of data, $\mathbf{X}$, for some interesting input feature, $X_j$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ with regard to a single instance of an input feature, $x_j$, and for it's corresponding single instance $\mathbf{x}$. PD and ICE can be overlaid in the same plot to create a holistic global and local portrait of the predictions for some $g$ and $X_j$ \cite{ice_plots}. Combining PD and ICE can also be indicative of modeled interactions in $g$ or of known weaknesses in PD estimation (e.g., inaccuracy in the presence of strong interactions and correlations \cite{ice_plots}, \cite{ale_plot}) when $\text{PD}(X_j, g)$ and $\text{ICE}(x_j, \mathbf{x}, g)$ diverge. For details regarding the calculation of one-dimensional PD and ICE, see the software resources in Subsection \ref{ssec:soft} and Appendices \ref{a_ssec:not} and \ref{a_ssec:pd_ice}.

@@

Shapley explanations are a class of additive, locally accurate feature contribution measures with long-standing theoretical support \cite{shapley}, \cite{shapley1988shapley}. Shapley explanations are the only possible locally accurate and globally consistent feature contribution values, meaning that Shapley explanation values for input features always sum to $g(\mathbf{x})$ for some $\mathbf{x} \in \mathcal{X}$ and that Shapley explanation values should never decrease in magnitude for some $x_j$ when $g$ is changed such that $x_j$ truly makes a stronger contribution to $g(\mathbf{x})$ \cite{shapley}, \cite{tree_shap}.

Shapley values can be estimated in different ways, many of which are intractable for datasets with large numbers of input features. Tree SHAP is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP (SHapley Additive exPlanations) has been implemented efficiently in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in Sections \ref{sec:m_and_m} and \ref{sec:res}. Deep SHAP is an approximate Shapley value technique that creates SHAP values for ANNs \cite{shapley}. Deep SHAP is implemented in the \href{https://github.com/slundberg/shap}{\texttt{shap}} package and is used to generate SHAP values for the two $g^{XNN}$ models discussed in Sections \ref{sec:m_and_m} and \ref{sec:res}. For more information pertaining to the calculation of Shapley values, see Appendices \ref{a_ssec:not} and \ref{a_ssec:shap}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\subsection{Discrimination Testing Metrics}\label{ssec:di}

Because many technical and academic discussions of fairness in ML have been inconclusive\footnote{See: \href{https://www.youtube.com/watch?v=jIXIuYdnyyk}{Tutorial: 21 Fairness Definitions and Their Politics}.}, this text will draw on regulatory and legal standards that have been used for years in industries like Financial Services. The discussed metrics are also representative of fair lending analyses and pair well with the mortgage data. See Appendix \ref{a_sec:types} for a brief discussion regarding different types of discrimination under US laws and Appendix \ref {a_sec:prac_v_stat} for remarks on statistical vs. practical significance for discrimination measure values.

One measure is known as marginal effects (ME). ME is simply the difference between the percent of the control group members receiving a favorable outcome and the percent of the protected class members receiving a favorable outcome.

@@

\noindent where $X_p$ and $X_c$ represent binary markers created from some demographic attribute, $c$ denotes the control group (often whites or males), and $p$ indicates a protected group.  ME is a favored DI metric used by the CFPB, the primary agency charged with regulating fair lending laws at the largest US lending institutions and various other participants in the consumer financial market.\footnote{See: \href{https://files.consumerfinance.gov/f/201510_cfpb_supervisory-highlights.pdf}{Supervisory Highlights, Issue 9, Fall 2015}.} Another important measure of DI is adverse impact ratio (AIR), more commonly known as a \textit{relative risk ratio} in settings outside of regulatory compliance.

@@

\noindent AIR is equal to the ratio of the proportion of the protected class that receives a favorable outcome divided by the proportion of the control class that receives a favorable outcome.  Another long-standing measure of DI is standardized mean difference (SMD). SMD is often used to assess disparities in continuous features, such as income differences in employment analyses, or interest rate differences in lending. It originated from work on statistical power, and is more formally known as \textit{Cohen’s d}. SMD is equal to the difference in the average class outcomes minus the control class outcome, divided by a measure of the standard deviation of the population.\footnote{There are several measures of the standard deviation of the score that are typically used: \textbf{1.} the standard deviation of the population, irrespective of protected class status, \textbf{2.} a standard deviation calculated only over the two groups being considered in a particular calculation, or \textbf{3.} a pooled standard deviation, using the standard deviations for each of the two groups with weights.} Cohen defined values of this metric to have \textit{small}, \textit{medium}, and \textit{large} effect sizes if the values exceeded 0.2, 0.5, and 0.8, respectively.

@@

\noindent The numerator in the SMD is equivalent to marginal effects but adds the standard deviation divisor as a standardizing factor.  Because of this standardization factor, SMD allows for a comparison across different types of outcomes, such as inequity in mortgage closing fees or inequities in the interest rates given on certain loans.  In this, one may apply definitions in \citet{cohen1988statistical} of \textit{small}, \textit{medium}, and \textit{large} effect sizes, which represent a measure of \textit{practical significance}, which is described in detail below.  Finally, confusion matrix metrics and their ratios in demographic groups are also considered as measures of DI in section \ref{sec:res}.

@@

@@

Results for the mortgage data are presented in Subsections \ref{ssec:c_v_uc_mort} -- \ref{ssec:dis_mort}.  $g^\text{ANN}$ and $g^\text{XNN}$ outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data, but as in Subsection \ref{ssec: c_v_uc_sim} the constrained variants of both model architectures do not show large differences in model performance with respect to unconstrained variants. Assuming that small fit differences on static test data do not outweigh the need for intrinsic model interpretability and reliable post-hoc explainability in high-stakes, human-centered, or regulated applications, only $g^\text{MGBM}$ and $g^\text{XNN}$ interpretability, post-hoc explainability, and discrimination testing results are presented.

@@

Following \citet{esl} a single feature $X_j \in \mathbf{X}$ and its complement set $\mathbf{X}_{\mathcal{P} \setminus \{j\}} \in \mathbf{X}$ (where $X_j \cup \mathbf{X}_{\mathcal{P} \setminus \{j\}} = \mathbf{X}$) is considered. $\text{PD}(X_j, g)$ for a given feature $X_j$ is estimated as the average output of the learned function $g(\mathbf{X})$ when all the observations of $X_j$ are set to a constant $x \in \mathcal{X}$ and $\mathbf{X}_{\mathcal{P} \setminus \{j\}}$ is left unchanged. $\text{ICE}(x_j, \mathbf{x}, g)$ for a given instance $\mathbf{x}$ and feature $x_j$ is estimated as the output of $g(\mathbf{x})$ when $x_j$ is set to a constant $x \in \mathcal{X}$ and all other features $\mathbf{x} \in \mathbf{X}_{\mathcal{P} \setminus \{j\}}$ are left untouched. PD and ICE curves are usually plotted over some set of constants $x \in \mathcal{X}$, as displayed in Section \ref{sec:res}. Due to known problems for PD in the presence of strong correlation and interactions, PD should not be used alone. PD should be paired with ICE or be replaced with accumulated local effect (ALE) plots \cite{ice_plots}, \cite{ale_plot}.

@@

\noindent In Equation \ref{eq:shap_additive}, $\mathbf{z} \in \{0,1\}^\mathcal{P}$ is a binary representation of $\mathbf{x}$ where 0 indicates missingness. Each $\phi_j$ is the local feature contribution value associated with $x_j$ and $\phi_0$ is the average of $g(\mathbf{X})$. Each $\phi_j$ is a weighted combination of model scores, $g_x(\mathbf{x})$, with $x_j$, $g_x(S \cup \{j\})$, and the model scores without $x_j$, $g_x(S)$, for every subset of features $S$ not including $j$, $S \subseteq \mathcal{P} \setminus \{j\}$, where $g_x$ incorporates the mapping between $\mathbf{x}$ and the binary vector $\mathbf{z}$.

@@

\section{Machine Learning Discrimination Under US Law}\label{a_sec:types}

Before discussing the techniques, it is important to explain and draw a distinction between the two major types of discrimination recognized in US legal and regulatory settings, disparate treatment (DT), and disparate impact (DI). DT (which is loosely referred to as \textit{intentional discrimination}) occurs most often in an algorithmic setting when a model explicitly uses protected class status (e.g., race, sex) as an input feature or uses a feature that is so similar to protected class status that it essentially proxies for class membership.  With some limited exceptions, the use of these factors in an algorithm is illegal under several statutes in the US.\textsuperscript{4} DI, colloquially known as \textit{unintentional discrimination}, occurs when some element of a decisioning process includes a \textit{facially neutral} factor (i.e, a reasonable and valid predictor of response) that results in a disproportionate share of a protected class receiving an unfavorable outcome.  In modeling, this is most typically driven by a statistically important feature that is distributed unevenly across classes, which causes more frequent unfavorable outcomes for the protected class.  However, other factors, such as hyperparameter or algorithm choices, can drive DI.  Crucially, legality hinges on whether changing the model, for example exchanging one feature for another or altering the hyperparameters of an algorithm, can lead to a similarly predictive model with lower disparate impact. The analyses and metrics herein focus on several measures of disparate impact that are commonly used in US litigation and regulatory settings.

\section{Practical vs. Statistical Significance for Discrimination Metrics}\label{a_sec:prac_v_stat}

A finding of \textit{practical significance} means the disparity found is not only statistically significant, but also passes beyond a chosen threshold that would constitute \textit{prima facia} evidence of illegal discrimination.  Its use represents a recognition that any large dataset is likely to show statistically significant differences in outcomes by class, even if those differences are not truly meaningful.  It further recognizes that there are likely to be situations where differences in outcomes are beyond a model user’s ability to correct them without significantly degrading the quality of the model. Moreover, practical significance is also needed by model builders and compliance personnel to determine whether a model should undergo remediation efforts before it is put into production. Unfortunately, guidelines for practical significance, i.e., the threshold at which any statistically significant disparity would be considered evidence of illegal discrimination, are not as frequently codified as the standards for statistical significance.  One exception, however, is in employment discrimination analyses, where the US Equal Employment Opportunity Commission (EEOC) has stated that if the AIR is below 0.80 and statistically significant, then this constitutes \textit{prima facia} evidence of discrimination, which the model user must rebut in order for the disparate impact not to be considered illegal discrimination.\footnote{Importantly, the standard of 0.80 is not a law, but a rule of thumb for agencies tasked with enforcement of discrimination laws.  ``Adoption of Questions and Answers To Clarify and Provide a Common Interpretation of the Uniform Guidelines on Employee Selection Procedures,'' Federal Register, Volume 44, Number 43 (1979).} It is important to note that the 0.80 measure of practical significance, also known as the \textit{80\% rule} and the \textit{4/5ths rule}, is explicitly used in relation to AIR, and it is not clear that the use of this threshold is directly relevant to testing fairness for metrics other than the AIR.

@@

\section{Simulated Data Results}

>>>>

\noindent where $\mathbf{X}_{\text{Friedman},j}$ are random uniform features in $[0,1]$. In Friedman's texts, a Gaussian noise term was added to create a continuous output feature for testing spline regression methodologies. In this manuscript, the signal generating function and input features are modified in several ways.  Two binary features, a categorical feature with five discrete levels, and a bias term are introduced into $f$ to add a degree of complexity that may more closely mimic real-world settings. 
%When generating the data, random draws for each observation are constructed such that the features can be correlated with each other, some highly so. For example, X1 and X3 have a Pearson correlation coefficient equal to 0.58.
For binary classification analysis, the Gaussian noise term is replaced with noise drawn from a logistic distribution and coefficients are re-scaled to be one fifth of the size of those used by Friedman, and any $f(\mathbf{X})$ value above 0 is classified as a positive outcome, while $f(\mathbf{X})$ values less than or equal to zero are designated as negative outcomes.  Finally, $f$ is augmented with two hypothetical protected class-control features with known dependencies on the binary outcome to allow for discrimination testing. The simulated data is generated to have eight input features, twelve after numeric encoding of categorical features, and a binary outcome, two class-control features, and 100,000 instances. The simulated data is then split into a training and test set, with 80,000 and 20,000 observations, respectively.  Within the training set, a 5-fold cross-validation indicator is used for training all models. For an exact specification of the simulated data, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints, with respect to each input feature, $X_j$, and a target feature, $\mathbf{y}$, independently. An MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, and each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$. For some instance, $\mathbf{x}$, a trained MGBM model, $g^{\text{MGBM}}$, takes the form:

@@

\noindent As in unconstrained GBM, $\Theta^{\text{mono}}_b$ is selected in a greedy, additive fashion by minimizing a regularized loss function that considers known target labels, the predictions of all subsequently trained trees in $g^{\text{MGBM}}$, and the $b$-th tree splits applied to $\mathbf{x}$, $T_b(\mathbf{x}^{(i)};\Theta^\text{mono}_b)$, in a numeric loss function (e.g., squared loss, Huber loss), and a regularization term that penalizes complexity in the current tree. See Appendices \ref{a_ssec:not} and \ref{a_ssec:mgbm} for details pertaining to MGBM training.

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bear a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e., GAMs which consider main effects and a small number of 2-way interactions and may also incorporate boosting into their training \cite{esl}, \cite{ga2m}.  Hence, XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. A trained XNN function, $g^{\text{XNN}}$, applied to some instance , $\mathbf{x}$, is defined as:

@@

Two $g^\text{XNN}$ models are trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in 5 training folds and in a test data partition. $L_1$ regularization is applied to all network weights to induce a sparse and interpretable model, where each $n_k$ and corresponding $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft} and Appendices  \ref{a_ssec:not} and \ref{a_ssec:xnn}.

@@

PD plots are a widely-used method for describing and plotting the estimated average prediction of a complex model, $g$, across some partition of data, $\mathbf{X}$, for some interesting input feature, $X_j$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ with regard to values of an input feature, $x_j$, in a single instance, $\mathbf{x}$. PD and ICE can be overlaid in the same plot to create a holistic global and local portrait of the predictions for some $g$ and $X_j$ \cite{ice_plots}. When $\text{PD}(X_j, g)$ and $\text{ICE}(x_j,  g)$ curves diverge, such plots can also be indicative of modeled interactions in $g$ or expose flaws in PD estimation, e.g., inaccuracy in the presence of strong interactions and correlations \cite{ice_plots}, \cite{ale_plot}.  For details regarding the calculation of one-dimensional PD and ICE, see the software resources in Subsection \ref{ssec:soft} and Appendices \ref{a_ssec:not} and \ref{a_ssec:pd_ice}.

@@

Shapley explanations are a class of additive, locally accurate feature contribution measures with long-standing theoretical support \cite{shapley}, \cite{shapley1988shapley}. Shapley explanations are the only locally accurate and globally consistent feature contribution values, meaning that Shapley explanation values for input features always sum to $g(\mathbf{x})$ for any $\mathbf{x}$, and that Shapley explanation values should not decrease in magnitude for some $x_j$ when $g$ is changed such that $x_j$ truly makes a stronger contribution to $g(\mathbf{x})$ \cite{shapley}, \cite{tree_shap}. Shapley values can be estimated in different ways, many of which are intractable for datasets with large numbers of input features. Tree SHAP (SHapley Additive exPlanations) is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP has been implemented in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in Section \ref{sec:res} and Appendix \ref{a_sec:res_sim}. Deep SHAP is an approximate Shapley value technique that creates SHAP values for ANNs \cite{shapley}. Deep SHAP is implemented in the \href{https://github.com/slundberg/shap}{\texttt{shap}} package and is used to generate SHAP values for the two $g^{XNN}$ models discussed in Section \ref{sec:res} and Appendix \ref{a_sec:res_sim}. For more information pertaining to the calculation of Shapley values, see Appendices \ref{a_ssec:not} and \ref{a_ssec:shap}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\subsection{Discrimination Testing Measures}\label{ssec:di}

Because many technical and academic discussions of fairness in ML have been inconclusive\footnote{See: \href{https://www.youtube.com/watch?v=jIXIuYdnyyk}{Tutorial: 21 Fairness Definitions and Their Politics}.}, this text will draw on regulatory and legal standards that have been used for years in regulated, high-stakes employment and financial decisions. The discussed measures are also representative of fair lending analyses and pair well with the mortgage data. See Appendix \ref{a_sec:types} for a brief discussion regarding different types of discrimination under US laws and Appendix \ref{a_sec:prac_v_stat} for remarks on statistical vs. practical significance for discrimination measures. One common measure of disparate impact used in US litigation and regulatory settings is known as marginal effects (ME). ME is simply the difference between the percent of the control group members receiving a favorable outcome and the percent of the protected class members receiving a favorable outcome.

@@

\noindent where $X_p$ and $X_c$ represent binary markers created from some demographic attribute, $c$ denotes the control group (often whites or males), $p$ indicates a protected group, and $\text{Pr}(\cdot)$ is the operator for conditional probability.  ME is a favored DI metric used by the the US Consumer Finance Protection Bureau (CFPB), the primary agency charged with regulating fair lending laws at the largest US lending institutions and for various other participants in the consumer financial market.\footnote{See: \href{https://files.consumerfinance.gov/f/201510_cfpb_supervisory-highlights.pdf}{Supervisory Highlights, Issue 9, Fall 2015}.} Another important DI measure is adverse impact ratio (AIR), more commonly known as a \textit{relative risk ratio} in settings outside of regulatory compliance.

@@

\noindent AIR is equal to the ratio of the proportion of the protected class that receives a favorable outcome divided by the proportion of the control class that receives a favorable outcome. Statistically significant AIR values below 0.8 can be considered \textit{prima facia} evidence of discrimination. An additional long-standing measure of DI is standardized mean difference (SMD). SMD is often used to assess disparities in continuous features, such as income differences in employment analyses, or interest rate differences in lending. It originates from work on statistical power, and is more formally known as \textit{Cohen’s d}. SMD is equal to the difference in the average protected class outcome, $\bar{\hat{\mathbf{y}}}_p$, minus the control class outcome, $\bar{\hat{\mathbf{y}}}_c$, divided by a measure of the standard deviation of the population, $\sigma_{\hat{\mathbf{y}}}$.\footnote{There are several measures of the standard deviation of the score that are typically used: \textbf{1.} the standard deviation of the population, irrespective of protected class status, \textbf{2.} a standard deviation calculated only over the two groups being considered in a particular calculation, or \textbf{3.} a pooled standard deviation, using the standard deviations for each of the two groups with weights.} Cohen defined values of this metric to have \textit{small}, \textit{medium}, and \textit{large} effect sizes if the values exceeded 0.2, 0.5, and 0.8, respectively.

@@

\noindent The numerator in the SMD is equivalent to marginal effects but adds the standard deviation divisor as a standardizing factor.  Because of this standardization factor, SMD allows for a comparison across different types of outcomes, such as inequity in mortgage closing fees or inequities in the interest rates given on certain loans.  In this, one may apply definitions in \citet{cohen1988statistical} of \textit{small}, \textit{medium}, and \textit{large} effect sizes, which represent a measure of \textit{practical significance}, which is described in detail in Appendix \ref{a_sec:prac_v_stat}. Finally, confusion matrix measures in demographic groups, such as accuracy, false negative rate (FNR), and their ratios, are also considered as measures of DI in Section \ref{sec:res} and Appendix \ref{a_sec:res_sim}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulated Data Results}

As model training and assessment on the simulated data is a rough validation exercise meant to showcase expected results on data with known characteristics, and given that most of the techniques in the proposed workflow are already used widely or have been validated elsewhere, reporting of simulated data results in the main text will focus mostly on debunking the perceived \textit{accuracy-interpretability tradeoff} for the considered model architectures and on the more novel XNN results. The bulk of the post-hoc explanation and discrimination testing results for the simulated are left to Appendix \ref{a_sec:res_sim}.

\subsubsection{Constrained vs. Unconstrained Model Fit Assessment}\label{ssec:c_v_uc_sim}

Model fit is roughly uniform for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data. Interpretability and explainability benefits of the constrained models appear to come at little cost to overall model performance, or in the case of $g^\text{ANN}$ and $g^\text{XNN}$, no cost at all. $g^\text{XNN}$ actually shows slightly better fit than $g^\text{ANN}$ across accuracy, area under the curve (AUC), logloss, and root mean squared error (RMSE). Accuracy is measured at the best F1 threshold for each model. Table \ref{tab:sim_acc} presents a variety of fit metrics for the $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data. $g^\text{GBM}$ exhibits the best performance, but all models give relatively similar fit results. Interpretability and explainability benefits of the constrained models appear to come at little cost to overall model performance, or in the case of $g^\text{ANN}$ and $g^\text{XNN}$, no cost at all. $g^\text{XNN}$ actually shows slightly better fit than $g^\text{ANN}$ across accuracy, area under the curve (AUC), logloss, and root mean squared error (RMSE). Accuracy is measured at the best F1 threshold for each model.

% add error metrics

\subsubsection{$g^\text{XNN}$ Interpretability Results}

For $g^\text{XNN}$, inherent interpretability manifests as plots of sparse $\gamma_k$ output layer weights, $n_k$ subnetwork ridge functions, and sparse $\beta_j$ weights in the projection layer. Figure \ref{fig:sim_xnn_glob_ridge} provides detailed insights into $g^\text{XNN}$. \ref{fig:sim_xnn_glob_ridge} \textbf{A} displays the sparse $\gamma_k$ weights of the output layer, where only $n_k$ subnetworks with $k \in \{1, 4, 7, 8, 9\}$ are associated with large magnitude weights. The $n_k$ subnetwork ridge functions appear in \ref{fig:sim_xnn_glob_ridge} \textbf{B} as simplistic but distinctive functional forms. Color-coding between \ref{fig:sim_xnn_glob_ridge} \textbf{A} and \ref{fig:sim_xnn_glob_ridge} \textbf{B} visually reinforces the direct feed-forward relationship between the $n_k$ subnetworks and the $\gamma_k$ weights of the output layer.

\begin{figure}[H]
	\hskip-20pt\includegraphics[width=18cm]{img/sim_xnn_glob_ridge.png}
	\caption{\textbf{A.} Output layer $\gamma_k$ weights, \textbf{B.} corresponding $n_k$ ridge functions, and \textbf{C.} associated projection layer $\beta_j$ weights for $g^\text{XNN}$ on the simulated test data.}
	\label{fig:sim_xnn_glob_ridge}
\end{figure}

% update figure

@@

Results for the mortgage data are presented in Subsections \ref{ssec:c_v_uc_mort} -- \ref{ssec:dis_mort} to showcase the example workflow.  $g^\text{ANN}$ and $g^\text{XNN}$ outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data, but as in Subsection \ref{ssec: c_v_uc_sim} the constrained variants of both model architectures do not show large differences in model performance with respect to unconstrained variants. Assuming that small fit differences on static test data do not outweigh the need for intrinsic model interpretability and reliable post-hoc explainability in high-stakes, human-centered, or regulated applications, only $g^\text{MGBM}$ and $g^\text{XNN}$ interpretability, post-hoc explainability, and discrimination testing results are presented.

@@

Following \citet{esl} a single feature $X_j \in \mathbf{X}$ and its complement set $\mathbf{X}_{\mathcal{P} \setminus \{j\}} \in \mathbf{X}$ (where $X_j \cup \mathbf{X}_{\mathcal{P} \setminus \{j\}} = \mathbf{X}$) is considered. $\text{PD}(X_j, g)$ for a given feature $X_j$ is estimated as the average output of the learned function $g(\mathbf{X})$ when all the observations of $X_j$ are set to a constant $x \in \mathcal{X}$ and $\mathbf{X}_{\mathcal{P} \setminus \{j\}}$ is left unchanged. $\text{ICE}(x_j, g)$ for a given instance $\mathbf{x}$ and feature $x_j$ is estimated as the output of $g(\mathbf{x})$ when $x_j$ is set to a constant $x \in \mathcal{X}$ and all other features $\mathbf{x} \in \mathbf{X}_{\mathcal{P} \setminus \{j\}}$ are left untouched. PD and ICE curves are usually plotted over some set of constants $x \in \mathcal{X}$, as displayed in Section \ref{sec:res}. Due to known problems for PD in the presence of strong correlation and interactions, PD should not be used alone. PD should be paired with ICE or be replaced with accumulated local effect (ALE) plots \cite{ice_plots}, \cite{ale_plot}.

@@

\noindent In Equation \ref{eq:shap_additive}, $\mathbf{z} \in \{0,1\}^\mathcal{P}$ is a binary representation of $\mathbf{x}$ where 0 indicates missingness. Each $\phi_j$ is the local feature contribution value associated with $x_j$, and $\phi_0$ is the average of $g(\mathbf{X})$. Each $\phi_j$ is a weighted combination of model scores, $g_x(\mathbf{x})$, with $x_j$, $g_x(S \cup \{j\})$, and the model scores without $x_j$, $g_x(S)$, for every subset of features $S$ not including $j$, $S \subseteq \mathcal{P} \setminus \{j\}$, where $g_x$ incorporates the mapping between $\mathbf{x}$ and the binary vector $\mathbf{z}$.

@@

\section{Machine Learning and Discrimination in US Legal and Regulatory Settings}\label{a_sec:types}

It is important to explain and draw a distinction between the two major types of discrimination recognized in US legal and regulatory settings, disparate treatment (DT), and disparate impact (DI). DT (which is loosely referred to as \textit{intentional discrimination}) occurs most often in an algorithmic setting when a model explicitly uses protected class status (e.g., race, sex) as an input feature or uses a feature that is so similar to protected class status that it essentially proxies for class membership.  With some limited exceptions, the use of these factors in an algorithm is illegal under several statutes in the US.\textsuperscript{4} DI, colloquially known as \textit{unintentional discrimination}, occurs when some element of a decisioning process includes a \textit{facially neutral} factor (i.e, a reasonable and valid predictor of response) that results in a disproportionate share of a protected class receiving an unfavorable outcome.  In modeling, this is most typically driven by a statistically important feature that is distributed unevenly across classes, which causes more frequent unfavorable outcomes for the protected class.  However, other factors, such as hyperparameter or algorithm choices, can drive DI.  Crucially, legality hinges on whether changing the model, for example exchanging one feature for another or altering the hyperparameters of an algorithm, can lead to a similarly predictive model with lower disparate impact.

\section{Practical vs. Statistical Significance for Discrimination Measures}\label{a_sec:prac_v_stat}

A finding of \textit{practical significance} means that discovered disparity is not only statistically significant, but also passes beyond a chosen threshold that would constitute \textit{prima facia} evidence of illegal discrimination.  Practical significance acknowledges that any large dataset is likely to show statistically significant differences in outcomes by class, even if those differences are not truly meaningful.  It further recognizes that there are likely to be situations where differences in outcomes are beyond a model user’s ability to correct them without significantly degrading the quality of the model. Moreover, practical significance is also needed by model builders and compliance personnel to determine whether a model should undergo remediation efforts before it is put into production. Unfortunately, guidelines for practical significance, i.e., the threshold at which any statistically significant disparity would be considered evidence of illegal discrimination, are not as frequently codified as the standards for statistical significance.  One exception, however, is in employment discrimination analyses, where the US Equal Employment Opportunity Commission (EEOC) has stated that if the AIR is below 0.80 and statistically significant, then this constitutes \textit{prima facia} evidence of discrimination, which the model user must rebut in order for the disparate impact not to be considered illegal discrimination.\footnote{Importantly, the standard of 0.80 is not a law, but a rule of thumb for agencies tasked with enforcement of discrimination laws.  ``Adoption of Questions and Answers To Clarify and Provide a Common Interpretation of the Uniform Guidelines on Employee Selection Procedures,'' Federal Register, Volume 44, Number 43 (1979).} It is important to note that the 0.80 measure of practical significance, also known as the \textit{80\% rule} and the \textit{4/5ths rule}, is explicitly used in relation to AIR, and it is not clear that the use of this threshold is directly relevant to testing fairness for measures other than the AIR.

@@

\section{Additional Simulated Data Results}\label{a_sec:res_sim}

Given that little or no trade-off is required in terms of model to fit to use the constrained models, intrinsic interpretability, post-hoc explainability, and discrimination are explored further for the $g^\text{MGBM}$ and $g^\text{XNN}$ models in Sections \ref{ssec:int_ex_sim} - \ref{ssec:dis_sim}. For $g^\text{MGBM}$, intrinsic interpretability is evaluated with PD and ICE plots of mostly monotonic prediction behavior for several important $X_j$, and post-hoc Shapley explanation analysis is used to create global and local feature importance. For $g^\text{XNN}$, post-hoc Shapley explanation techniques are also used to generate global and local feature importance to augment interpretability results exhibited in Section \ref{}. Both $g^\text{MGBM}$ and $g^\text{XNN}$ are evaluated for discrimination using AIR, ME, SMD, and other measures.
